Computational biology in the 21st century:\\Scaling with compressive algorithms

### 2015-09-22 main.tex
Put quartz first

Compressing reads for storage is necessary should researchers wish to apply more advanced mapping tools or other analysis in the future.
Taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common short---15bp---substring) together into `buckets', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This approach allows a general-purpose compressor to achieve better compression.
SCALCE~\cite{berger2013computational} also relies on a `boosting' scheme, 
reordering
reads in such a way that a general-purpose compressor achieves improved
compression.

@@

The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST (compressively-accelerated BLAST)~\cite{loh2012compressive} and 
caBLASTP~\cite{daniels2013compressive}, accelerates existing tools (in this 
case, read
mappers including BWA or Bowtie2) by allowing them to operate in a compressed
space, and relies on a coarse and a fine phase.
In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{berger2013computational}, conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as `seeds' for longer sequence matches.
CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each 
one to its neighbors within a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.
As an aside, though not an example of accelerated similarity search, mid-size $l$-mers
can also be used for storage compression of NGS quality scores associated with base calls.
Unlike many other approaches to compressing quality scores in the literature
~\cite{bonfield2013compression,ochoa2013qualcomp},
Quartz~\cite{yu2015quality} takes advantage of the fact that mid-size $l$-mers can in many cases almost uniquely identify locations in the genome,
bounding the likelihood that a quality score is informative and allowing for lossy compression of uninformative scores.
Because Quartz's lossy compression injects information from the distribution of $l$-mers in the target genome, it demonstrates not only improved compression over competing
approaches, but slightly improves the accuracy of downstream 
variant-calling~\cite{yu2015quality}.

>>>>

Compressing reads for storage is necessary should researchers wish to apply more advanced mapping tools or other analysis in the future \cite{bonfield2013compression}.
As stated earlier, NGS reads consist of a sequence string and associated quality scores, the latter of which generally uses more space when compressed.
By taking advantage of biological structure, both parts of NGS reads can be better compressed.
Unlike some other approaches to compressing quality scores in the literature
~\cite{bonfield2013compression,ochoa2013qualcomp},
Quartz~\cite{yu2015quality} takes advantage of the fact that mid-size $l$-mers can in many cases almost uniquely identify locations in the genome,
bounding the likelihood that a quality score is informative and allowing for lossy compression of uninformative scores.
approaches, but slightly improves the accuracy of downstream 
variant-calling~\cite{yu2015quality}.
Because Quartz's lossy compression injects information from the distribution of $l$-mers in the target genome, it demonstrates not only improved compression over competing approaches, but slightly improves the accuracy of downstream variant-calling \cite{yu2015quality}.
Similarly, for the sequence string,
taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common short---15bp---substring) together into `buckets', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This approach allows a general-purpose compressor to achieve better compression.
SCALCE~\cite{berger2013computational} also relies on a `boosting' scheme, 
reordering
reads in such a way that a general-purpose compressor achieves improved
compression.

@@

The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST (compressively-accelerated BLAST)~\cite{loh2012compressive} and 
caBLASTP~\cite{daniels2013compressive}, accelerates existing tools (in this 
case, read
mappers including BWA or Bowtie2) by allowing them to operate in a compressed
space, and relies on a coarse and a fine phase.
In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{berger2013computational}, conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as `seeds' for longer sequence matches.
CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each 
one to its neighbors within a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.
